Linking watershed nutrient loading to estuary water quality with Generalized Additive Models

### 2023-08-08 Schramm-2023-06-PeerJ/Schramm-2023-08-PeerJ.tex
reorder by estuary location

Water quality data for the two freshwater and three estuary locations
were obtained from the Texas Commission on Environmental Quality (TCEQ)
Surface Water Quality Monitoring Information System. Data submitted
through the system are required to be collected under Quality Assurance
Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. These operating procedures ensure consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying time periods and availability of water quality data. For
freshwater locations, TP from January 2000 through December 2020 and
nitrate-nitrogen (NO\textsubscript{3}) data from January 2005 through
December 2020 were downloaded (Table~\ref{tab:table1}). Less than
5-years of total nitrogen and TKN concentration data were available at
the freshwater sites and deemed insufficient to develop load estimation
models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper Lavaca Bay site near the
outlet of the Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site
(TCEQ-13383), and the lower Lavaca Bay site near the mouth of the Bay
(TCEQ-13384). For estuary locations, we obtained data for TP,
Nitrite+Nitrate (NO\textsubscript{\emph{X}}), TKN, chlorophyll-\emph{a},
and DO concentrations from January 2005 through December 2020
(Table~\ref{tab:table2}).

@@

\caption{\label{tab:table2}Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}llll>{\raggedright\arraybackslash}p{2.5cm}rr}
\toprule
Station ID & Parameter & Mean & SD & N & Method & AWRL\textsuperscript{a} & Criteria\textsuperscript{b}\\
\midrule
 & TP (mg/L) & 0.11 & 0.05 & 47 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13383, Upper-Bay} & DO (mg/L) & 7.22 & 1.35 & 55 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.08 & 0.03 & 51 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13384, Lower-Bay} & DO (mg/L) & 7.51 & 1.32 & 54 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.13 & 0.06 & 50 & EPA 365.1 & 0.06 & 0.21\\

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13563, Mid-Bay} & DO (mg/L) & 7.91 & 1.34 & 56 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item[a] Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.
\item[b] The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes. NO\textsubscript{x} and TKN parameters are not currently screened by the state and do not have an associated screening level.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

Moving average (\emph{ma}) is an exponentially smoothed moving average
that attempts to incorporate the influence of prior streamflow events on
concentration at the current time period
\autocite{wangLoadEstimationUncertainties2011,kuhnertQuantifyingTotalSuspended2012,zhang_improving_2017},
using Eq. \ref{eq2}:

\begin{equation}
ma(t) = \frac{\sum^{J}_{j=1}{d^{t-t_j}Q_{t_{j}}}}{\sum^{J}_{j=1}d^{t-t_j}}
\label{eq2}
\end{equation}

where \emph{t} is the day of estimation, \emph{d} is the discount factor
(set to 0.95), \emph{J} is the total number of daily discharge
observations, \(t-t_j\) is the lag time between the current day and some
historical day of observation (\(t_j\)). Here, as \emph{d} approaches
zero, the time series becomes the daily observed values and as \emph{d}
gets closer to one, the time series becomes the mean flow.

Flow anomaly (\emph{fa}) is a unitless term that represents how wet or
dry the current time period is from a previous time period
\autocite{vecchiaTrendsConcentrationsUse2009,zhang_improving_2017}.
Long-term flow anomaly (\emph{ltfa}) is the streamflow over the previous
year relative to the entire period
\autocite[Eq.~\ref{eq3},][]{zhang_improving_2017} and and the short-term
flow anomaly (\emph{stfa}) calculated as the current day flow compared
to the preceding 1-month streamflow
\autocite[Eq.~\ref{eq4},][]{zhang_improving_2017}:

@@

where \(\mu\) is the conditional expected response (nutrient
concentration), \(g()\) is the log link, and response variable was
modeled as Gamma distributed with mean \(\mu\) and scale \(\lambda\).
\(f_1(ddate)\) is decimal date smoothed with a thin-plate regression
spline, \(f_2(yday)\) is the numeric day of year smoothed with a cyclic
cubic regression spline. Prior work has shown that many water quality
parameters may have lagged effects lasting days or even months following
storms and large discharge events
\autocite{mooneyWatershedExportEvents2012a,wetzExtremeFutureEstuaries2013,bukaveckasInfluenceStormEvents2020,walkerTimescalesMagnitudeWater2021}.
\textcite{walkerTimescalesMagnitudeWater2021} showed the impact of large
flow events like Hurricane Harvey had somewhat short water quality
impacts in Lavaca Bay and adjacent estuary systems, ranging somewhere
around 20 days to a few months. We incorporate this lag effect by using
cumulative totals for \(f_3(Q)\) and \(f_4(Load)\). \(f_3(Q)\) is the
cumulative of the previous 20 days of inflow (combined measurements from
Lavaca River and Navidad River) and \(f_4(Load)\) is the cumulative of
the previous 20 days total NO\textsubscript{3} or TP watershed load. The
set of models specified for each water quality response are in
Table~\ref{tab:table3}.

@@

\caption{Density plots of goodness-of-fit metrics (NSE, \textit{r}, and PBIAS) from repeated 5-fold cross validation between predicted nutrient loads from GAM models and measured nutrient loads. Color indicates the tail probability calculated from the empirical cumulative distribution of the goodness-of-fit metrics.}\label{fig:fig2}
\end{figure}

@@

There was no evidence of long-term changes in TP or DO concentrations at
any Lavaca Bay site (Fig.~\ref{fig:fig5}). The upper-bay site,
TCEQ-13563, had evidence of a long-term linear increase in
NO\textsubscript{\emph{X}} while chlorophyll-\emph{a} decreased from
2005 through 2014 (Fig.~\ref{fig:fig5}). NO\textsubscript{\emph{X}}
concentration at the mid-bay site, TCEQ-13383, displayed an unusual
periodic pattern that is indicative of a strong influence from inflow or
precipitation. The temporal GAMs did not provide evidence of long-term
trends in any of the water quality constituents at the lower-bay
TCEQ-13384 site.

@@

\caption{Fitted splines (shaded regions indicate 90\% confidence intervals) from the temporal estuary GAM display the marginal smoothed effect of date on TP (A), NO\textit{\textsubscript{X}} (B), chlorophyll-\textit{a} (C), TKN (D), and DO (E) concentrations at each site in Lavaca Bay.}\label{fig:fig5}
\end{figure}

@@

\caption{\label{tab:table4}Estuary GAM AIC\textsubscript{c} values and associated model probabilities (in parenthesis). Models with the highest probability for each site and water quality parameter combination are bolded and italicized for emphasis.}
\centering
\begin{tabular}[t]{ll>{}l>{}l>{}l}
\toprule
Parameter & Site & Temporal & Inflow & Inflow + Load\\
\midrule
 & Upper-Bay & -152.1 (0.02) & -155.1 (0.07) & \em{\textbf{-160.2 (0.91)}}\\

 & Mid-Bay & -145.3 (0.01) & -150 (0.12) & \em{\textbf{-154 (0.87)}}\\

\multirow{-3}{*}{\raggedright\arraybackslash TP} & Lower-Bay & -194.4 (0.39) & \em{\textbf{-194 (0.31)}} & -194 (0.31)\\
\cmidrule{1-5}
 & Upper-Bay & -218.9 (0) & \em{\textbf{-241.7 (0.5)}} & -241.7 (0.5)\\

 & Mid-Bay & -175.1 (0.01) & \em{\textbf{-183.8 (0.5)}} & -183.8 (0.5)\\

\multirow{-3}{*}{\raggedright\arraybackslash NO\textsubscript{x}} & Lower-Bay & -263.4 (0) & -298.1 (0.5) & \em{\textbf{-298.1 (0.5)}}\\
\cmidrule{1-5}
 & Upper-Bay & 279.7 (0.24) & \em{\textbf{279.6 (0.26)}} & 278.2 (0.5)\\

 & Mid-Bay & 289.5 (0.37) & \em{\textbf{289.5 (0.38)}} & 290.3 (0.25)\\

\multirow{-3}{*}{\raggedright\arraybackslash Chlorophyll-\textit{a}} & Lower-Bay & \em{\textbf{268.2 (0.03)}} & 262.7 (0.48) & 262.7 (0.48)\\
\cmidrule{1-5}
 & Upper-Bay & \em{\textbf{42.2 (0.5)}} & 42.2 (0.5) & -\\

 & Mid-Bay & 31.1 (0.5) & \em{\textbf{31.1 (0.5)}} & -\\

\multirow{-3}{*}{\raggedright\arraybackslash TKN} & Lower-Bay & \em{\textbf{34.3 (0.5)}} & 34.3 (0.5) & -\\
\cmidrule{1-5}
 & Upper-Bay & \em{\textbf{146.4 (0.37)}} & 146.8 (0.29) & 146.5 (0.34)\\

 & Mid-Bay & 138.3 (0.17) & \em{\textbf{136.4 (0.42)}} & 136.4 (0.42)\\

\multirow{-3}{*}{\raggedright\arraybackslash DO} & Lower-Bay & \em{\textbf{135.9 (0.04)}} & 130.6 (0.48) & 130.6 (0.48)\\
\bottomrule
\end{tabular}
\end{table}

@@

TP and NO\textsubscript{3} loadings from the Lavaca Bay watershed showed
high inter-annual variability driven primarily by fluctuations in
discharge. Notably, there were no indications of trends in
flow-normalized NO\textsubscript{3} and TP loads in the Navidad River.
In comparison, there was weak evidence of more recent decreases in
flow-normalized NO\textsubscript{3} (but not TP) loads in the Lavaca
River watershed. While the dominant agricultural land uses differ
between the Lavaca (primarily grazed pasture and rangeland) and Navidad
(mix of pasture and row crops) catchments, we did not have a reason to
expect different flow normalized trends between the two systems from
land use alone. Freshwater discharges in the Navidad River are regulated
by the Palmetto Bend Dam forming Lake Texana at the lower extent of the
river. Lentic nitrogen uptake and cycling may have regulating effects
that mask changes in upstream nitrogen loadings. Additional nutrient
data collection in the contributing tributaries of Lake Texana is needed
to fully assess the role of Lake Texana in regulating nutrient delivery
to the Lavaca Bay system. However, these results suggest that there have
been no changes in the NO\textsubscript{3} or TP loading from the
Navidad River system at the Lake Texana discharge point when accounting
for variations in year to year discharge.

@@

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{X}}, TKN, and TP concentration are suggestive of
nonpoint watershed sources, consistent with watershed uses and with
other studies relating freshwater inflow with nutrient concentrations in
Lavaca Bay and other estuaries
\autocite{russell_effect_2006,caffreyHighNutrientPulses2007,peierlsNonmonotonicResponsesPhytoplankton2012,palmerImpactsDroughtsLow2015,ciraPhytoplanktonDynamicsLowinflow2021}.
Inflow had a non-linear relationship with TKN at the two upstream sites,
with TKN increasing as freshwater inflow transitioned from low to
moderate levels. At higher freshwater inflows, the effect was
attenuated, possibly indicating a flushing effect at higher freshwater
inflow. No relationship between TKN and freshwater inflow were observed
at TCEQ-13384 located in the lower reach of Lavaca Bay. The results
suggest that tidal flushing might dilute TKN and act as a control on the
effects of freshwater inflow in the lower reaches of Lavaca Bay. The
results are also consistent with previous work that suggest processing
of organic loads in the upper Lavaca Bay or tidal portions of the Lavaca
River reduce transport of nutrients to the lower reaches of Lavaca Bay
\autocite{russell_effect_2006}.

@@

The GAM approach proved useful for both estimating loads and assessing
downstream responses in water quality. Although we did not compare other
models, it is likely similar estimates of loadings would be obtained by
methods suggest as LOADEST, WRTDS, or SPARROW given the functionally
similar dependent variable structures. The underlying weakness in the
estimates of loading in the current study is the reliance on ambient
water quality data used for statewide water quality assessments.
Cross-validation of the nutrient loading models highlights that
predictions are prone to high bias, owing to the lack of targeted storm
or flow biased measurements. The high biases are indicative that subsets
of values were unable to capture the functional relationships with the
flow based dependent variables. It was beyond the scope of the current
study to evaluate the subsets of cross-validation data and scores.
However, the cross-validation procedure is indicative that more robust
sampling is needed. Supplementary flow-biased monitoring targeting
storm- or high-flow conditions is critical to improve model performance
and strength of evidence produced by these models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
Although there is existing work on the samples sizes and sample design
required for reliable performance of both LOADEST
\autocite{parkUsePollutantLoad2014} and WRTDS
\autocite{kumarValueIntensiveSampling2019} models, similar work does not
appear to have been extended to water quality applications of GAMs.

>>>>

Water quality data for the two freshwater and three estuary locations
were obtained from the Texas Commission on Environmental Quality (TCEQ)
Surface Water Quality Monitoring Information System. Data submitted
through the system are required to be collected under Quality Assurance
Project Plans and lab method procedures outlined by the TCEQ's
procedures manual. These operating procedures ensure consistent
collection and laboratory methods are applied between samples collected
by different entities and under different projects. All sites had
varying time periods and availability of water quality data. For
freshwater locations, TP from January 2000 through December 2020 and
nitrate-nitrogen (NO\textsubscript{3}) data from January 2005 through
December 2020 were downloaded (Table~\ref{tab:table1}). Less than
5-years of total nitrogen and TKN concentration data were available at
the freshwater sites and deemed insufficient to develop load estimation
models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
The three estuary sites included an upper-bay site near the outlet of
the Lavaca River system (TCEQ-13563), a mid-bay site (TCEQ-13383), and
the lower-bay site near the mouth of the Bay (TCEQ-13384). For estuary
locations, we obtained data for TP, Nitrite+Nitrate
(NO\textsubscript{\emph{X}}), TKN, chlorophyll-\emph{a}, and DO
concentrations from January 2005 through December 2020
(Table~\ref{tab:table2}). All water quality data was collected
approximately quarterly.

@@

\caption{\label{tab:table2}Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}llll>{\raggedright\arraybackslash}p{2.5cm}rr}
\toprule
Station ID & Parameter & Mean & SD & N & Method & AWRL\textsuperscript{a} & Criteria\textsuperscript{b}\\
\midrule
 & TP (mg/L) & 0.13 & 0.06 & 50 & EPA 365.1 & 0.06 & 0.21\\

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13563, Upper-Bay} & DO (mg/L) & 7.91 & 1.34 & 56 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.11 & 0.05 & 47 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13383, Mid-Bay} & DO (mg/L) & 7.22 & 1.35 & 55 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\cmidrule{1-8}
 & TP (mg/L) & 0.08 & 0.03 & 51 & EPA 365.1 & 0.06 & 0.21\\

@@

\multirow{-5}{2cm}{\raggedright\arraybackslash TCEQ-13384, Lower-Bay} & DO (mg/L) & 7.51 & 1.32 & 54 & ASTM D888-09(C) and TCEQ SOP V1 &  & 5.00\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item[a] Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.
\item[b] The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes. NO\textsubscript{x} and TKN parameters are not currently screened by the state and do not have an associated screening level.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

Moving average (\emph{ma}) is an exponentially smoothed moving average
that attempts to incorporate the influence of prior streamflow events on
concentration at the current time period
\autocite{wangLoadEstimationUncertainties2011,kuhnertQuantifyingTotalSuspended2012,wangSedimentConcentrationPrediction2013a,zhang_improving_2017},
using Eq. \ref{eq2}:

\begin{equation}
ma(d) = \frac{Q_j + dQ_{j-1} + \mathellipsis + d^{j-1}Q_1}{1+d+...+d^{j-1}}
\label{eq2}
\end{equation}

where \emph{ma} at discount factor \emph{d} is calculated using flows
\((Q_1,Q_2,...,Q_j)\) for days 1 through \emph{j}. Here, as \emph{d}
approaches zero, the time series becomes the daily observed values and
as \emph{d} gets closer to one, the time series becomes the mean flow.
Although \textcite{kuhnertQuantifyingTotalSuspended2012} and
\textcite{wangSedimentConcentrationPrediction2013a} suggest using
multiple covariates with \emph{d} set at different values from 0.75 to
0.99, \textcite{robson_prediction_2015-1} and
\textcite{zhang_improving_2017} found substantial improvements in
statistical nutrient models with \emph{d} equal to 0.95 which was
adopted in this study.

Flow anomaly (\emph{fa}) is a dimensionless term that represents how wet
or dry (or how anomalous) the current time period is from previous time
periods
\autocite{vecchiaTrendsConcentrationsUse2009,zhang_improving_2017}.
Long-term flow anomaly (\emph{ltfa}) is the streamflow over the previous
year relative to the entire period
\autocite[Eq.~\ref{eq3},][]{zhang_improving_2017} and and the short-term
flow anomaly (\emph{stfa}) calculated as the current day flow compared
to the preceding 1-month streamflow
\autocite[Eq.~\ref{eq4},][]{zhang_improving_2017}:

@@

where \(\mu\) is the conditional expected response (nutrient
concentration), \(g()\) is the log link, and response variable was
modeled as Gamma distributed with mean \(\mu\) and scale \(\lambda\).
\(f_1(ddate)\) is decimal date smoothed with a thin-plate regression
spline, \(f_2(yday)\) is the numeric day of year smoothed with a cyclic
cubic regression spline. Prior work has shown that many water quality
parameters may have lagged effects lasting days or even months following
storms and large discharge events
\autocite{mooneyWatershedExportEvents2012a,wetzExtremeFutureEstuaries2013,bukaveckasInfluenceStormEvents2020,walkerTimescalesMagnitudeWater2021}.
\textcite{walkerTimescalesMagnitudeWater2021} showed the impact of large
flow events like Hurricane Harvey had somewhat short water quality
impacts in Lavaca Bay and adjacent estuary systems, ranging somewhere
around 20 days to a few months. We incorporate this lag effect by using
cumulative totals for \(f_3(Q)\) and \(f_4(Load)\). \(f_3(Q)\) is the
cumulative of the previous 20 days of inflow (combined measurements from
Lavaca River and Navidad River) and \(f_4(Load)\) is the cumulative of
the previous 20 days total NO\textsubscript{3} or TP watershed load,
both smoothed with thin plate regression splines. The set of models
specified for each water quality response are in Table~\ref{tab:table3}.

@@

\caption{Density plots of goodness-of-fit metrics (NSE, \textit{r}, and PBIAS) from repeated 5-fold cross validation between predicted nutrient loads from GAM models and measured nutrient loads. Color indicates the tail probability calculated from the empirical cumulative distribution of the goodness-of-fit metrics. Values closer to 1 for NSE and r and values closer to 0 for PBIAS represent more ideal goodness-of-fit assessments.}\label{fig:fig2}
\end{figure}

@@

There was no evidence of long-term changes in TP or DO concentrations at
any Lavaca Bay site (Fig.~\ref{fig:fig5}). The upper-bay site,
TCEQ-13563, had evidence of a long-term linear increase in
NO\textsubscript{\emph{X}} while chlorophyll-\emph{a} decreased from
2005 through 2014 (Fig.~\ref{fig:fig5}). NO\textsubscript{\emph{X}}
concentration at the mid-bay site, TCEQ-13383, displayed a periodic
pattern that is indicative of a strong influence from inflow or
precipitation. The temporal GAMs did not provide evidence of long-term
trends in any of the water quality constituents at the lower-bay
TCEQ-13384 site.

@@

\caption{Fitted splines (shaded regions indicate 90\% confidence intervals) from the temporal estuary GAM (Table \ref{tab:table3}) display the marginal smoothed effect of date on TP (A), NO\textit{\textsubscript{X}} (B), chlorophyll-\textit{a} (C), TKN (D), and DO (E) concentrations at each site in Lavaca Bay.}\label{fig:fig5}
\end{figure}

@@

\caption{\label{tab:table4}Estuary GAM AIC\textsubscript{c} values and associated model probabilities (in parenthesis). Models with the highest probability for each site and water quality parameter combination are bolded and italicized for emphasis.}
\centering
\begin{tabular}[t]{ll>{}l>{}l>{}l}
\toprule
Parameter & Site & Temporal & Inflow & Inflow + Load\\
\midrule
 & Upper-Bay & -145.3 (0.01) & -150 (0.12) & \em{\textbf{-154 (0.87)}}\\

 & Mid-Bay & -152.1 (0.02) & -155.1 (0.07) & \em{\textbf{-160.2 (0.91)}}\\

\multirow{-3}{*}{\raggedright\arraybackslash TP} & Lower-Bay & \em{\textbf{-194.4 (0.39)}} & -194 (0.31) & -194 (0.31)\\
\cmidrule{1-5}
 & Upper-Bay & -175.1 (0.01) & \em{\textbf{-183.8 (0.5)}} & -183.8 (0.5)\\

 & Mid-Bay & -218.9 (0) & \em{\textbf{-241.7 (0.5)}} & -241.7 (0.5)\\

\multirow{-3}{*}{\raggedright\arraybackslash NO\textsubscript{x}} & Lower-Bay & -263.4 (0) & \em{\textbf{-298.1 (0.5)}} & -298.1 (0.5)\\
\cmidrule{1-5}
 & Upper-Bay & 289.5 (0.37) & \em{\textbf{289.5 (0.38)}} & 290.3 (0.25)\\

 & Mid-Bay & 279.7 (0.24) & 279.6 (0.26) & \em{\textbf{278.2 (0.5)}}\\

\multirow{-3}{*}{\raggedright\arraybackslash Chlorophyll-\textit{a}} & Lower-Bay & 268.2 (0.03) & \em{\textbf{262.7 (0.48)}} & 262.7 (0.48)\\
\cmidrule{1-5}
 & Upper-Bay & \em{\textbf{31.1 (0.5)}} & 31.1 (0.5) & -\\

 & Mid-Bay & \em{\textbf{42.2 (0.5)}} & 42.2 (0.5) & -\\

\multirow{-3}{*}{\raggedright\arraybackslash TKN} & Lower-Bay & \em{\textbf{34.3 (0.5)}} & 34.3 (0.5) & -\\
\cmidrule{1-5}
 & Upper-Bay & 138.3 (0.17) & \em{\textbf{136.4 (0.42)}} & 136.4 (0.42)\\

 & Mid-Bay & \em{\textbf{146.4 (0.37)}} & 146.8 (0.29) & 146.5 (0.34)\\

\multirow{-3}{*}{\raggedright\arraybackslash DO} & Lower-Bay & 135.9 (0.04) & \em{\textbf{130.6 (0.48)}} & 130.6 (0.48)\\
\bottomrule
\end{tabular}
\end{table}

@@

TP and NO\textsubscript{3} loadings from the Lavaca Bay watershed showed
high inter-annual variability driven primarily by fluctuations in
discharge. Notably, there were no indications of trends in
flow-normalized NO\textsubscript{3} and TP loads in the Navidad River.
In comparison, there was weak evidence of more recent decreases in
flow-normalized NO\textsubscript{3} (but not TP) loads in the Lavaca
River watershed. While the dominant agricultural land uses differ
between the Lavaca (primarily grazed pasture and rangeland) and Navidad
(mix of pasture and row crops) catchments, we did not have a reason to
expect different flow normalized trends between the two systems from
land use alone. Freshwater discharges in the Navidad River are regulated
by the Palmetto Bend Dam forming Lake Texana at the lower extent of the
river. Lentic nitrogen uptake and cycling may have regulating effects
that mask changes in upstream nitrogen loadings. Additional nutrient
data collection in the tributaries of Lake Texana is needed to fully
assess the role of Lake Texana in regulating nutrient delivery to the
Lavaca Bay system. However, these results suggest that there have been
no changes in the NO\textsubscript{3} or TP loading from the Navidad
River system at the Lake Texana discharge point when accounting for
variations in year to year discharge.

@@

The strong positive effect of freshwater inflow on
NO\textsubscript{\emph{X}}, TKN, and TP concentration are suggestive of
nonpoint watershed sources, consistent with watershed uses and with
other studies relating freshwater inflow with nutrient concentrations in
Lavaca Bay and other estuaries
\autocite{russell_effect_2006,caffreyHighNutrientPulses2007,peierlsNonmonotonicResponsesPhytoplankton2012,palmerImpactsDroughtsLow2015,ciraPhytoplanktonDynamicsLowinflow2021}.
Inflow had a non-linear relationship with TKN at the two upstream sites,
with TKN increasing as freshwater inflow transitioned from low to
moderate levels. At higher freshwater inflows, the effect was
attenuated, possibly indicating a flushing effect at higher freshwater
inflow. No relationship between TKN and freshwater inflow was observed
at TCEQ-13384 located in the lower reach of Lavaca Bay. The results
suggest that tidal flushing might dilute TKN and act as a control on the
effects of freshwater inflow in the lower reaches of Lavaca Bay. The
results are also consistent with previous work that suggest processing
of organic loads in the upper Lavaca Bay or tidal portions of the Lavaca
River reduce transport of nutrients to the lower reaches of Lavaca Bay
\autocite{russell_effect_2006}.

@@

The GAM approach proved useful for both estimating loads and assessing
downstream responses in water quality. Although we did not compare other
models, it is likely similar estimates of loadings would be obtained by
methods such as LOADEST, WRTDS, or SPARROW given the functionally
similar dependent variable structures. The underlying weakness in the
estimates of loading in the current study is the reliance on ambient
water quality data used for statewide water quality assessments.
Cross-validation of the nutrient loading models highlights that
predictions are prone to high bias, owing to the lack of targeted storm
or flow biased measurements. The high biases are indicative that subsets
of values were unable to capture the functional relationships with the
flow based dependent variables. It was beyond the scope of the current
study to evaluate the subsets of cross-validation data and scores.
However, the cross-validation procedure is indicative that more robust
sampling is needed. Supplementary flow-biased monitoring targeting
storm- or high-flow conditions is critical to improve model performance
and strength of evidence produced by these models
\autocite{horowitzEvaluationSedimentRating2003,snelderEstimationCatchmentNutrient2017}.
Although there is existing work on the samples sizes and sample design
required for reliable performance of both LOADEST
\autocite{parkUsePollutantLoad2014} and WRTDS
\autocite{kumarValueIntensiveSampling2019} models, similar work does not
appear to have been extended to water quality applications of GAMs.
